Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-05-30 pituba/article/article_spe.tex
hotel attempt

\begin{document}
\begin{frontmatter}
\title{Solution of hydodynamic equilibrium in black-oil reservoirs: the classical material balance equation}
%\title{Obtaining the classical material balance equation from black-oil equations}
%\journal{Journal of Petroleum Science and Engineering}
%\author{R.B. Piccinini}\ead{rbpiccinini@gmail.com}
%\address{Petrobras - UO-BA/ATP-N/RES. Salvador, Bahia, Brazil.}

@@

\begin{abstract}
This work derives the classical material balance equation from a three dimensional black-oil model. Sufficient conditions for solution uniqueness of material balance equation are presented. It is shown that spatially averaged PVT properties are required for reducing a black-oil model to the material balance approach, it is shown, though, that the averaging process may be neglected for many practical purposes.  An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined. A scheme numerically solve material balance equation for pressure is also proposed.
%
%This work derives the classical material balance equation from a black-oil model. We examine the existing link between both approaches in detail, including the necessary step of averaging PVT properties. An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. A scheme for numerical solution is also proposed.
\end{abstract}

@@

The term black-oil refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure may be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component may be defined for a simplified description. A pseudo oil component may also be defined by subtracting the pseudo gas component from oil phase composition. After defining these two pseudo-components, the number of transport equations needed for hydrocarbon species is reduced from a usually large number to only two.

@@

The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied to material balance. This approach extends material balance equation to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, depends on how important are non-equilibrium effects to phase equilibrium.

@@

%\section{Black-oil equations}
%
%Here it is used a thermodynamic fluid model known as Black-Oil \cite{blackoil}. It consists of a simplified treatment of hydrocarbon composition and phase equilibrium. Hydrocarbon species are grouped into two pseudo-components called oil and gas, being a pseudo-component one which consists of more than one chemical molecule.
%
%Oil is composed by hydrocarbon species that remain in liquid phase for standard surface pressure and temperature. The other hydrocarbon species in gaseous phase for the same surface conditions are called gas.
%
%The mixture of reservoir fluid is then composed of oil, gas and water components and can present three fluid phases: oil, gas and water. Oil phase may be composed of oil and gas components, gaseous phase may be composed uniquely by gas component and water phase uniquely by water component.
%
%In addition to the three fluid components and phases, there is a solid phase composed by the rock.
%
%The fluid phases may be defined in a more general framework so as to account for dissolution of gas in water or water vapour, for instance. Such possibilities are not dealt with in this work.
%
%For some volume containing one of the three fluid phases, the saturation of the fluid phase $k$ is defined as
%\begin{equation}\label{eq: saturation}
%S_{k}=\frac{\phi_k}{\sum_{i \neq solid} \phi_i} \, ,
%\end{equation}
%where $\phi_i$ is the volume fraction of the fluid phase $i$.
%
%In order to apply the volume balance, a control volume ($\Omega$) is defined containing all the hydrocarbon zone. It consists of the region where hydrocarbon saturation is greater than zero for some arbitrarily small volume. This reservoir definition excludes the pure water zone, which is referred to as the aquifer.
%
%
%Right hand side of Equation \eqref{eq: matbal1}, \textit{rock and fluid expansion}, is a function of phase mass transfers and compressibilities. Both should be obtained from a thermodynamic model accounting for mass transfer among phases and the relation between phase volumes and pressure. Temperature is seldom considered to vary during reservoir production and it is held constant and equal in all phases, not being an independent property of the control volume state.
%
%Such thermodynamic model is presented in detail in \cite{blackoil} and it is commonly referred to as Black-Oil Equations. The main assumption behind them is that phase and pressure equilibria are instantaneously achieved inside the control volume, i.e., mass transfer among phases reaches equilibrium and internal forces equilibrate external ones fast enough in comparison to the interval of time between two observations of control volume properties.
%
%Pressure in each phase may differ if capillary forces are taken into account, for what extra constitutive equations relating each phase pressures would be needed.
%Capillary pressure would be function of porous medium properties given by some porous medium model, e.g., permeability and porosity; phase properties, e.g., surface tension and wettability, and the thermodynamic state.
%
%Capillary pressure is very often assumed to be function of phase saturations only and, hence, the thermodynamic state of the control volume would be completely determined by the pressure of one phase (e.g. the wetting phase) and phase saturations.
%
%Black-Oil Equations do not assume water and oil phases becoming miscible for any thermodynamic state.
%
%Capillary pressure dependence on phase saturation only is often considered as a weak assumption in contradiction with experimental observations. The hysteresis of capillary pressure exhibited when either imbibition or drainage occurs is not theoretically explained and it is accounted for by empirical relations or it is rather simply neglected.
%
%\cite{hilfer1} has proposed the inclusion of an energy equation.

@@

% Oil component, also called stock tank oil, is defined as the hydrocarbon composition in liquid phase for standard surface pressure and temperature. Similarly, gas component is the hydrocarbon composition in gaseous phase for the same surface conditions. All species are generally present in both phases although the equilibrium composition is shifted to liquid phase (or gaseous phase) for components of greater (lower) molecular mass.

Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases: oil, gas and liquid water. Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases.

In addition to the three fluid components and phases, there is the solid phase composed by the rock.

@@

\item gas component:
\begin{equation}\label{eq: Sg1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV + \oint_{\partial \Omega} \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) \cdot \vec{dS}= \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}
Mass concentration of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively.

Phase saturation is defined as its volumetric fraction of the pore space:
\begin{equation}
S_\alpha = \frac{\phi_\alpha}{\phi} \, ,
\end{equation}
where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity.
\begin{equation}
S_o + S_g +S_w = 1 \, .
\end{equation}

A common simplification in black-oil formulations is to restrict mass transport to macroscopic advection. No diffusive or dispersive term is included in the mass balance equations.

@@

Another simplification is now made assuming thermal equilibrium in the entire reservoir volume:
\begin{equation}
T_w = T_o = T_g = T_{rock} = T
\end{equation}
and
\begin{equation}
\nabla T \equiv 0 \, .
\end{equation}

Temperature is thus a constant prescribed value and no energy balance equation is needed. It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components. As temperature is assumed constant, oil phase density in saturated states may be a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase.

@@

We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at some given pressure and temperature condition to the volume of the same phase at standard pressure and temperature, i.e.:

@@

In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between volumes of dissolved gas and oil both measured in standard condition:

@@

Equations may now be rewritten using the above definitions and denoting porous volume by $V_p = \bar{\phi}\Omega$.

@@

Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at standard conditions and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured in reservoir conditions.

An equation for bubble-point may be obtained by setting $S_g=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure $R_s(p=P_b)$:
\begin{equation}
\begin{split}
R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
&+G_i-G_p\Big]/\left(N-N_p\right) \, .
\end{split}
\end{equation}

@@

Eq~\eqref{eq: MBE} only involves fluid properties, original fluid volumes ($N$, $W$ and $G)$ and cumulative production and injection.
It is an agelbraic equation that may be solved to any of its terms provided all remaining are known. Several techniques exist in the literature to estimate original volume of fluids, water influx from aquifers or reservoir pressure.

\subsubsection{Definition of a reference depth for pressure}

Material balance equation is widely used to stablish a relationship between production (and injection) and reservoir pressure. In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties,  which depend on pressure. It turns out, though, that pressure is a function of depth and phase.

@@

@@

For the case of an oil reservoir covered by a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact. The point here is that at phase contact both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value (we are neglecting capillary pressure).

@@

For a non-volatile saturated oil, we may approximate $dR_s / dp_o$ to the ratio between  solubility ratio ($R_{s}$) and saturation pressure ($p_{sat}$), so that Eq.~\eqref{eq: dRsdz1} becomes:
\begin{equation}\label{eq: dRsdz2}
\frac{1}{R_s}\frac{d R_s}{dz} \approx \frac{\rho_o g }{p_{sat}} \approx 10^{-3}\ m^{-1} \, ,
\end{equation}
where the following orders of magnitudes were assumed: $p_{sat} \approx 10^{7}\ Pa$, $\rho_o \approx 10^{3}\ kg/m^3$ and $g \approx 10\ m^2/s$.

@@

\section*{Nomenclature}
\begin{tabular}{ll}
$B_g$ & gas formation volume factor\\
$B_o$ & oil formation volume factor\\
$B_w$ & water formation volume factor\\
$R_s$ & solubility ratio of gas in oil phase\\
$S_g$ & gas saturation \\
$S_o$ & oil saturation \\
$S_w$ & water saturation \\
$G_p$ & produced volume of gas \\
$m$ & ratio of original free gas volume to original oil phase volume in reservoir conditions \\
$N$ & original volume of oil in place \\
$N_p$ & produced volume of oil \\
$P_b$ & bubble-point pressure \\
$V_p$ & porous volume \\
$W_i$ & injected volume of water \\
$W_p$ & produced volume of water
\end{tabular}
\section*{References}
\bibliographystyle{elsarticle-harv}
\bibliography{refs}
\cite{*}

>>>>

\begin{document}
\begin{frontmatter}
\title{Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation}
%\title{Obtaining the classical material balance equation from black-oil equations}
%\journal{Journal of Petroleum Science and Engineering}
%\author{R.B. Piccinini}\ead{rbpiccinini@gmail.com}
%\address{Petrobras - UO-BA/ATP-N/RES. Salvador, Bahia, Brazil.}

@@

\begin{abstract}
This work derives the classical material balance equation from a three dimensional black-oil model. Sufficient conditions for solution uniqueness of material balance equation are presented for the case that pressure is the unknown quantity. It is shown that spatially averaged PVT properties are required for reducing a black-oil model to the material balance approach. It is also shown that the averaging process may be neglected for many practical purposes.  An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined. A scheme numerically solve material balance equation for pressure is also proposed.
%
%This work derives the classical material balance equation from a black-oil model. We examine the existing link between both approaches in detail, including the necessary step of averaging PVT properties. An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. A scheme for numerical solution is also proposed.
\end{abstract}

@@

The term black-oil refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can be defined for a simplified description gas phase composition. A pseudo oil component can also be defined by subtracting the pseudo gas component from oil phase composition. After defining these two pseudo-components, the number of transport equations needed for hydrocarbon species is reduced from usually a large number to only two.

@@

The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied here to material balance. This approach extends material balance equation to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, depends on how important non-equilibrium effects are to phase equilibrium.

@@

%\section{Black-oil equations}
%
%Here it is used a thermodynamic fluid model known as Black-Oil \cite{blackoil}. It consists of a simplified treatment of hydrocarbon composition and phase equilibrium. Hydrocarbon species are grouped into two pseudo-components called oil and gas, being a pseudo-component one which consists of more than one chemical molecule.
%
%Oil is composed by hydrocarbon species that remain in liquid phase for standard surface pressure and temperature. The other hydrocarbon species in gaseous phase for the same surface condition are called gas.
%
%The mixture of reservoir fluid is then composed of oil, gas and water components and can present three fluid phases: oil, gas and water. Oil phase may be composed of oil and gas components, gaseous phase may be composed uniquely by gas component and water phase uniquely by water component.
%
%In addition to the three fluid components and phases, there is a solid phase composed by the rock.
%
%The fluid phases may be defined in a more general framework so as to account for dissolution of gas in water or water vapour, for instance. Such possibilities are not dealt with in this work.
%
%For some volume containing one of the three fluid phases, the saturation of the fluid phase $k$ is defined as
%\begin{equation}\label{eq: saturation}
%S_{k}=\frac{\phi_k}{\sum_{i \neq solid} \phi_i} \, ,
%\end{equation}
%where $\phi_i$ is the volume fraction of the fluid phase $i$.
%
%In order to apply the volume balance, a control volume ($\Omega$) is defined containing all the hydrocarbon zone. It consists of the region where hydrocarbon saturation is greater than zero for some arbitrarily small volume. This reservoir definition excludes the pure water zone, which is referred to as the aquifer.
%
%
%Right hand side of Equation \eqref{eq: matbal1}, \textit{rock and fluid expansion}, is a function of phase mass transfers and compressibilities. Both should be obtained from a thermodynamic model accounting for mass transfer among phases and the relation between phase volumes and pressure. Temperature is seldom considered to vary during reservoir production and it is held constant and equal in all phases, not being an independent property of the control volume state.
%
%Such thermodynamic model is presented in detail in \cite{blackoil} and it is commonly referred to as Black-Oil Equations. The main assumption behind them is that phase and pressure equilibria are instantaneously achieved inside the control volume, i.e., mass transfer among phases reaches equilibrium and internal forces equilibrate external ones fast enough in comparison to the interval of time between two observations of control volume properties.
%
%Pressure in each phase may differ if capillary forces are taken into account, for what extra constitutive equations relating each phase pressures would be needed.
%Capillary pressure would be function of porous medium properties given by some porous medium model, e.g., permeability and porosity; phase properties, e.g., surface tension and wettability, and the thermodynamic state.
%
%Capillary pressure is very often assumed to be function of phase saturations only and, hence, the thermodynamic state of the control volume would be completely determined by the pressure of one phase (e.g. the wetting phase) and phase saturations.
%
%Black-Oil Equations do not assume water and oil phases becoming miscible for any thermodynamic state.
%
%Capillary pressure dependence on phase saturation only is often considered as a weak assumption in contradiction with experimental observations. The hysteresis of capillary pressure exhibited when either imbibition or drainage occurs is not theoretically explained and it is accounted for by empirical relations or it is rather simply neglected.
%
%\cite{hilfer1} has proposed the inclusion of an energy equation.

@@

% Oil component, also called stock tank oil, is defined as the hydrocarbon composition in liquid phase for standard surface pressure and temperature. Similarly, gas component is the hydrocarbon composition in gaseous phase for the same surface condition. All species are generally present in both phases although the equilibrium composition is shifted to liquid phase (or gaseous phase) for components of greater (lower) molecular mass.

Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases: oil, gas and (liquid) water. Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases.

%In addition to the three fluid components and phases, there is the solid phase composed by the rock.

@@

\item gas component:
\begin{equation}\label{eq: Sg1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV + \oint_{\partial \Omega} \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) \cdot \vec{dS}= \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}
Mass concentrations of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively.

Phase saturation is defined as its volumetric fraction of the pore space:
\begin{equation}
S_\alpha = \frac{\phi_\alpha}{\phi} \, ,
\end{equation}
where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity. Clearly, all saturations sum up to unity. 
\begin{equation}
S_o + S_g +S_w = 1 \, .
\end{equation}

A common simplification in black-oil formulation is to restrict mass transport to macroscopic advection. No diffusive or dispersive term is included in mass balance equations.

@@

Another simplification is now made assuming thermal equilibrium in the entire reservoir volume, i.e.,
\begin{equation}
T_w = T_o = T_g = T_{rock} = T = \text{constant} \, .
\end{equation}

Temperature is a constant prescribed value and thus no equation for energy balance is needed. It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if component number is restricted to the two pseudo-components. As temperature is assumed constant, oil phase density is a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase:

@@

We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to the volume at surface condition, i.e.:

@@

In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between volumes of dissolved gas and oil both measured in surface condition:

@@

Equations can now be rewritten using the above definitions and denoting porous volume by $V_p = \bar{\phi}\Omega$.

@@

Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at surface condition and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured in reservoir condition.

An equation for bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure $R_s(p=P_b)$:
\begin{equation}
\begin{split}
R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
&+G_i-G_p\Big]/\left(N-N_p\right) \, .
\end{split}
\end{equation}

@@

Eq~\eqref{eq: MBE} involves fluid properties, original fluid volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
It is an agelbraic equation that can be solved to any of its terms provided all remaining are known. This works focuses on its solution for pressure.

\subsubsection{An interpretation for calculated pressure and the definition of a reference depth}

Material balance equation is widely used to stablish a relationship between production (and injection) and reservoir pressure. In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties, and they are function of pressure. It turns out, though, that pressure is a function of depth and phase.

@@

Oil and gas densities are thus rather constant with depth and a zeroth order approximation of $\bar{\rho}$ is accurate within $0.5\%$ for material balance calculations.

@@

For the case of an oil reservoir covered by a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact. The point here is that at phase contact both gas and oil densities can be computed using the same pressure, i.e., the corresponding local value (we are neglecting capillary pressure).

@@

For a non-volatile saturated oil, we can approximate $dR_s / dp_o$ to the ratio between  solubility ratio ($R_{s}$) and saturation pressure ($p_{sat}$), so that Eq.~\eqref{eq: dRsdz1} becomes:
\begin{equation}\label{eq: dRsdz2}
\frac{1}{R_s}\frac{d R_s}{dz} \approx \frac{\rho_o g }{p_{sat}} \approx 10^{-3}\ m^{-1} \, ,
\end{equation}
where the following orders of magnitudes were assumed: $p_{sat} \approx 10^{7}\ Pa$, $\rho_o \approx 10^{3}\ kg/m^3$ and $g \approx 10\ m^2/s$.

@@

\section*{Nomenclature}
\begin{tabular}{ll}
$B_g$ & gas formation volume factor\\
$B_o$ & oil formation volume factor\\
$B_w$ & water formation volume factor\\
$R_s$ & solubility ratio of gas in oil phase\\
$S_g$ & gas saturation \\
$S_o$ & oil saturation \\
$S_w$ & water saturation \\
$G_p$ & produced volume of gas \\
$m$ & ratio of original free gas volume to original oil phase volume in reservoir condition \\
$N$ & original volume of oil in place \\
$N_p$ & produced volume of oil \\
$P_b$ & bubble-point pressure \\
$V_p$ & porous volume \\
$W_i$ & injected volume of water \\
$W_p$ & produced volume of water
\end{tabular}
\section*{References}
\bibliographystyle{elsarticle-harv}
\bibliography{refs}
\cite{*}
